Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-25 submission/manuscript.Rmd
make new edits and streamline results section

**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 2A]. Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 2B]. These observations provide evidence that gene prevalence does not change due to colorectal tumors.

**The number of significant positive correlations between OTU relative abundance and SCFA concentration are similar for individuals without tumors and patients with adenoma or carcinoma tumors.** Having found no difference in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeablely higher number of significant negative correlations associated with patients with adenoma versus the other groups for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family seemed to have the largest share of these negative correlations within patients with adenoma tumors [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3]. Overall, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace all important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F]. The reason for this seems to be because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C]. The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining [Figure 4B & 4C]. In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.

**Random Forest models for SCFA concentrations have different top 10 most important OTUs than disease models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high and low SCFA concentrations [Figure 5A & S2A]. However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [[Figure 5A & S2A]. Additionally, there were differences classification for both models based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B]. There was minimal overlap between these model's most important OTUs and those used to classify patients with either adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. The only OTUs that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. OTU00167 was in the top 10 most important OTUs for the OTU adenoma model and not in the SCFA and OTU adenoma model, however, acetate and butyrate concentrations were [Figure 4B-C]. These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate. Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with an adenoma or carcinoma tumor.

>>>>

**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 2A]. Using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 2B]. These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors.

**Total significant positive correlations between OTU relative abundance and SCFA concentration were similar for individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of significant OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeablely higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas  [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F]. The main reason for this was because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C]. The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining in the top 10 [Figure 4B & 4C]. In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.

**Random Forest models for SCFA concentrations have different top 10 most important OTUs than tumor models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher/lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high/low SCFA groups [Figure 5A & S2A]. However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure 5A & S2A]. Additionally, there were differences in accuracy for both model types based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B]. There also was minimal overlap between these SCFA classification model's most important OTUs and those used to classify patients with adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. The only OTU that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. Additionally, OTU00167 was in the top 10 most important OTUs for the OTU adenoma model but not in the SCFA and OTU adenoma model while acetate and butyrate concentrations were [Figure 4B-C]. These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate. Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with adenomas or carcinomas.

### 2018-07-25 submission/manuscript.tex
make new edits and streamline results section

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used high-performance liquid chromatography (HPLC) to measure
acetate, butyrate, and propionate concentrations of frozen fecal samples
from 490 individuals at a cross-sectional point in time. There was no
difference between individuals without colon tumors (n=172) and patients
with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
measured after multiple comparison correction (P-value \textgreater{}
0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
had pre- and post-treatment fecal samples. Although there was a general
trend for increasing acetate, butyrate, and propionate concentrations
after treatment for tumors, there was no significant difference pre- and
post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance between individuals without colon tumors
and patients with adenomas or carcinomas. First, using imputed gene
relative abundance based on 16S rRNA gene sequencing we found no
difference in any of the genes involved with acetate, butyrate, or
propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
similarity between groups is highlighted by visualizing genes important
in butyrate synthesis {[}Figure 2A{]}. Using a paired Wilcoxon rank-sum
test, there also was no difference in imputed gene relative abundance
between pre- and post-treatment samples for any genes involved with SCFA
synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
subset of these 490 fecal samples (n=85) and used metagenomic sequencing
to confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals without colon tumors (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 2B{]}. These
observations provide evidence that gene prevalence does not change due
to colorectal tumors.

\textbf{The number of significant positive correlations between OTU
relative abundance and SCFA concentration are similar for individuals
without tumors and patients with adenoma or carcinoma tumors.} Having
found no difference in SCFA concentrations or genes coding for enzymes
involved with SCFA synthesis, we next investigated if specific OTUs
correlated with SCFA concentrations. Using Spearmans rho we found that
the majority of OTU correlations were to taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} {[}Figure 3 \& Table
S5{]}. A similar pattern was observed when using high/low SCFA groups
based on the overall median concentration for that specific SCFA
{[}Figure S1 \& Table S6{]}. There was a noticeablely higher number of
significant negative correlations associated with patients with adenoma
versus the other groups for all SCFAs tested {[}Figure 3{]}. In
particular, OTUs from the \emph{Ruminococcaceae} family seemed to have
the largest share of these negative correlations within patients with
adenoma tumors {[}Figure 3{]}. The number of positive correlations
between OTUs and SCFA concentrations was similar between individuals
without tumors and patients with a carcinoma {[}Figure 3{]}. Although
patients with adenomas had more positive correlations between OTUs and
SCFA concentrations, their total number was more similar to the other
groups than the number of negative correlations {[}Figure 3{]}. Overall,
these results suggest that the resident taxa that may change the most
due to carcinoma tumors may not be ones that are responsible for the
production of acetate, butyrate, or propionate.

\textbf{SCFA concentrations do not replace all important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} SCFA
concentrations could improve prediction of tumors based on specific
bacterial community structures. Our OTU data can be used in combination
with SCFAs to assess whether there is a community dependent context to
SCFA classification of tumors. Using the Random Forest algorithm we
built models with OTU abundance data only or OTU abundances and SCFA
concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. For adenoma and carcinoma models, there was no
difference between the median AUC of models with or without SCFA
concentrations (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
There was little difference between the top 10 most important OTUs, as
measured by mean decrease in accuracy, in models using SCFA and OTUs
versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The reason for this seems to
be because SCFA concentrations replaced the information gained by
specific OTUs rather than add new information to the model {[}Figure
4B-C{]}. The SCFA concentrations also only replaced select OTUs with
taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} with many OTUs within
these taxa remaining {[}Figure 4B \& 4C{]}. In combination with the
previous results on taxa correlations, these observations provide
additional evidence that the resident taxa within models that classify
tumor are not ones associated with acetate, butyrate, or propionate
production.

\textbf{Random Forest models for SCFA concentrations have different top
10 most important OTUs than disease models.} Using OTU data we built
Random Forest models to classify either SCFA concentration or higher
than median and lower than median SCFA concentrations. Overall, OTU data
had a reasonable ability to classify both SCFA concentrations and high
and low SCFA concentrations {[}Figure 5A \& S2A{]}. However, these
models tended to be over fit, suggesting that rarer taxa may be
important for this classification {[}{[}Figure 5A \& S2A{]}.
Additionally, there were differences classification for both models
based on whether the individual had no tumors, an adenoma, or a
carcinoma {[}Figure 5B \& S2B{]}. There was minimal overlap between
these model's most important OTUs and those used to classify patients
with either adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and
S2C-E{]}. The only OTUs that did overlap between the models was OTU00167
(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
OTU00167 was in the top 10 most important OTUs for the OTU adenoma model
and not in the SCFA and OTU adenoma model, however, acetate and butyrate
concentrations were {[}Figure 4B-C{]}. These observations provide
further evidence that it is possible to identify specific OTUs
associated with higher SCFA concentrations and that these OTUs belong to
taxa known to produce acetate, butyrate, and propionate. Although it is
possible to identify OTUs associated with SCFA production, our results
do not support the hypothesis that SCFA concentration or OTUs associated
with their production are different between individuals with no tumors
and patients with an adenoma or carcinoma tumor.

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used high-performance liquid chromatography (HPLC) to measure
acetate, butyrate, and propionate concentrations of frozen fecal samples
from 490 individuals at a cross-sectional point in time. There was no
difference between individuals without colon tumors (n=172) and patients
with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
measured after multiple comparison correction (P-value \textgreater{}
0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
had pre- and post-treatment fecal samples. Although there was a general
trend for increasing acetate, butyrate, and propionate concentrations
after treatment for tumors, there was no significant difference pre- and
post-treatment for patients with adenomas (P-value \textgreater{} 0.20)
or carcinomas (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance between individuals without colon tumors
and patients with adenomas or carcinomas. First, using imputed gene
relative abundance based on 16S rRNA gene sequencing we found no
difference in any of the genes involved with acetate, butyrate, or
propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
similarity between groups is highlighted by visualizing genes important
in butyrate synthesis {[}Figure 2A{]}. Using a paired Wilcoxon rank-sum
test, there also was no difference in imputed gene relative abundance
between pre- and post-treatment samples for any genes involved with SCFA
synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
subset of these 490 fecal samples (n=85) and used metagenomic sequencing
to confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals without colon tumors (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference can be highlighted when we
visualize the results for butyrate kinase {[}Figure 2B{]}. These
observations provide evidence that gene prevalence for enzymes involved
in SCFA synthesis does not change due to colorectal tumors.

\textbf{Total significant positive correlations between OTU relative
abundance and SCFA concentration were similar for individuals without
tumors and patients with adenomas or carcinomas.} Having found no
difference between individuals without tumors and patients with adenomas
or carcinomas in SCFA concentrations or genes coding for enzymes
involved with SCFA synthesis, we next investigated if specific OTUs
correlated with SCFA concentrations. Using Spearmans rho we found that
the majority of significant OTU correlations were to taxa from
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
{[}Figure 3 \& Table S5{]}. A similar pattern was observed when using
high/low SCFA groups based on the overall median concentration for that
specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeablely
higher number of significant negative correlations associated with
patients with adenomas for all SCFAs tested {[}Figure 3{]}. In
particular, OTUs from the \emph{Ruminococcaceae} family had the largest
share of these negative correlations within patients with adenomas
{[}Figure 3{]}. Although patients with adenomas had more positive
correlations between OTUs and SCFA concentrations, their total number
was more similar to individuals without tumors or patients with
carcinomas versus the analogous comparison for the number of negative
correlations {[}Figure 3{]}. The number of positive correlations between
OTUs and SCFA concentrations was similar between individuals without
tumors and patients with a carcinoma {[}Figure 3{]}. Overall, these
results suggest that the resident taxa that may change the most due to
colon tumors may not be ones that are responsible for the production of
acetate, butyrate, or propionate.

\textbf{SCFA concentrations do not replace important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} SCFA
concentrations could improve prediction of tumors based on specific
bacterial community structures. Our OTU data can be used in combination
with SCFAs to assess whether there is a community dependent context to
SCFA classification of tumors. Using the Random Forest algorithm we
built models with OTU abundance data or OTU abundances and SCFA
concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. For adenoma and carcinoma models, there was no
difference between the median AUC of models with or without SCFA
concentrations (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
There was little difference between the top 10 most important OTUs, as
measured by mean decrease in accuracy, in models using SCFA and OTUs
versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The main reason for this was
because SCFA concentrations replaced the information gained by specific
OTUs rather than add new information to the model {[}Figure 4B-C{]}. The
SCFA concentrations also only replaced select OTUs with taxonomic
classification to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} with many OTUs within these taxa remaining in the
top 10 {[}Figure 4B \& 4C{]}. In combination with the previous results
on taxa correlations, these observations provide additional evidence
that the resident taxa within models that classify tumor are not ones
associated with acetate, butyrate, or propionate production.

\textbf{Random Forest models for SCFA concentrations have different top
10 most important OTUs than tumor models.} Using OTU data we built
Random Forest models to classify either SCFA concentration or
higher/lower than median SCFA concentrations. Overall, OTU data had a
reasonable ability to classify both SCFA concentrations and high/low
SCFA groups {[}Figure 5A \& S2A{]}. However, these models tended to be
over fit, suggesting that rarer taxa may be important for this
classification {[}Figure 5A \& S2A{]}. Additionally, there were
differences in accuracy for both model types based on whether the
individual had no tumors, an adenoma, or a carcinoma {[}Figure 5B \&
S2B{]}. There also was minimal overlap between these SCFA classification
model's most important OTUs and those used to classify patients with
adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
The only OTU that did overlap between the models was OTU00167
(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
Additionally, OTU00167 was in the top 10 most important OTUs for the OTU
adenoma model but not in the SCFA and OTU adenoma model while acetate
and butyrate concentrations were {[}Figure 4B-C{]}. These observations
provide further evidence that it is possible to identify specific OTUs
associated with higher SCFA concentrations and that these OTUs belong to
taxa known to produce acetate, butyrate, and propionate. Although it is
possible to identify OTUs associated with SCFA production, our results
do not support the hypothesis that SCFA concentration or OTUs associated
with their production are different between individuals with no tumors
and patients with adenomas or carcinomas.
